---
title: "Assignment 4"
author: "Mehnaz_meem"
date: "2025-10-05"
output:
  pdf_document: default
  html_document: default
  word_document: default
---

##The file “gene_expression.tsv” contains RNA-seq count data for three samples of interest.
##Read in the file, making the gene identifiers the row names. Show a table of values for the first six genes.

```{r}
library("R.utils")
```

```{r}
URL="https://raw.githubusercontent.com/ghazkha/Assessment4/refs/heads/main/gene_expression.tsv"
download.file(URL,destfile="gene_expression.tsv")
```
```{r}
list.files()
```

##The file “growth_data.csv” contains measurements for tree circumference growing at two sites
```{r}
URL="https://raw.githubusercontent.com/ghazkha/Assessment4/refs/heads/main/growth_data.csv"
download.file(URL,destfile="growth_data.csv")
```
```{r}
list.files()
```

##Question 1
##Read the file data, first six genes
```{r}
gene_data <- read.delim("gene_expression.tsv", row.names = 1, header = TRUE)
head(gene_data, 6)
```
##Question 2
##New column with mean of other columns

```{r}
gene_data$Mean <- rowMeans(gene_data)
```

##Showing table of values for first six genes
```{r}
head(gene_data, 6)
```
##Question 3
##List the 10 genes with the highest mean expression
```{r}
top10_genes <- gene_data[order(gene_data$Mean, decreasing = TRUE), ]
head(top10_genes, 10)
```

##Question 4
##Determine the number of genes with a mean <10
```{r}
num_genes_below_10 <- sum(gene_data$Mean < 10)
cat(num_genes_below_10)
```


##Question 5
##Make a histogram plot of the mean values
```{r}
hist(gene_data$Mean,
breaks = 1000,
main = "Distribution of Mean Gene Expression",
xlab = "Mean Expression Level",
ylab = "Frequency",
xlim = c(0,30000),
col = "blue")
```



##Question 6
##Import growth_data csv file into an R object
```{r}
growth <- read.csv("growth_data.csv", header = TRUE)
```

##Column names of growth_data
```{r}
colnames(growth)
```


##Question 7
##Calculate the mean and standard deviation of tree circumference at the start and end of the study at both sites
```{r}
# Mean and Standard Deviation (2005) & (2020)
growth_stats_raw <- aggregate(cbind(Circumf_2005_cm, Circumf_2020_cm)~ Site,
data = growth,
FUN = function(x) c(mean = mean(x, na.rm = TRUE),
sd = sd(x, na.rm = TRUE)))


# Split the matrix columns into separate columns
growth_stats_summary <- data.frame(
Site = growth_stats_raw$Site,
Mean_2005 = growth_stats_raw$Circumf_2005_cm[, "mean"],
SD_2005 = growth_stats_raw$Circumf_2005_cm[, "sd"],
Mean_2020 = growth_stats_raw$Circumf_2020_cm[, "mean"],
SD_2020 = growth_stats_raw$Circumf_2020_cm[, "sd"]
)



# Display table results
print(growth_stats_summary)
```

##Question 8
##Make a box plot of tree circumference at the start and end of the study at both sites.
```{r}
# Prepare data for boxplot
northeast_start <- growth$Circumf_2005_cm[growth$Site== "northeast"]
northeast_end <- growth$Circumf_2020_cm[growth$Site== "northeast"]
southwest_start <- growth$Circumf_2005_cm[growth$Site== "southwest"]
southwest_end <- growth$Circumf_2020_cm[growth$Site== "southwest"]


# Create boxplot
boxplot(northeast_start, northeast_end, southwest_start, southwest_end,
names = c("Northeast\nStart", "Northeast\nEnd", "Southwest\nStart", "Southwest\nEnd"),
main = "Tree Circumference by Site and Time Point",
xlab = "Site and Time Point",
ylab = "Circumference (cm)",
col = c("lightcoral", "blue", "lightcoral", "blue"),
border = "black")


# Add a legend
legend("topleft",
legend = c("Start", "End"),
fill = c("lightcoral", "blue"),
title = "Time Point")
```

##Question 9
##Calculate the mean growth over the last 10 years at each site.
```{r}
# Growth from 2010 to 2020
growth$Growth_10yr <- growth$Circumf_2020_cm- growth$Circumf_2010_cm


# Mean 10 year growth for each site
mean_growth_by_site <- aggregate(Growth_10yr~ Site, data = growth, FUN = mean)


# Show results
print(mean_growth_by_site)
```


##Question 10
##Using T-test to estimate the growth beteen two sites
```{r}
t_test_result <- t.test(Growth_10yr~ Site, data = growth)

## Print t-test results with interpretation
cat("\n T-Test: Comparing 10-Year Growth Between Sites\n")
```

```{r}
cat("--------------------------------------------------\n")
```

```{r}
cat("P-value:", round(t_test_result$p.value, 4), "\n")
```

```{r}
cat("95% Confidence Interval:",
round(t_test_result$conf.int[1], 2), "to",
round(t_test_result$conf.int[2], 2), "\n")
```

```{r}
cat("Mean Growth - Northeast:", round(t_test_result$estimate["mean in group Northeast"], 2), "cm\n")
```

```{r}
cat("Mean Growth - Southwest:", round(t_test_result$estimate["mean in group Southwest"], 2), "cm\n")
```

```{r}
# Corrected R Code for Interpretation Block
if (t_test_result$p.value < 0.05) {
  cat("Conclusion: There is a statistically significant difference in 10-year growth between sites (p < 0.05).\n")
} else {
  # Replacing the Unicode character (\u2265) with the ASCII text "> ="
  cat("Conclusion: No statistically significant difference in 10-year growth between sites (p >= 0.05).\n")
}
```

