---
title: "MEHNAZ_MEEM_A4_part_2"
author: "Mehnaz_meem"
date: "2025-10-06"
output:
  pdf_document: default
  html_document: default
  word_document: default
---


##Question 1
##Downloading the gemone


```{r}
library(R.utils)
```

```{r}
#Download and Process E. coli CDS
#E. coli K-12 MG1655 (GCA_000005845) CDS file
URL="http://ftp.ensemblgenomes.org/pub/bacteria/release-53/fasta/bacteria_0_collection/escherichia_coli_str_k_12_substr_mg1655_gca_000005845/cds/Escherichia_coli_str_k_12_substr_mg1655_gca_000005845.ASM584v2.cds.all.fa.gz"
download.file(URL,destfile="ecoli_cds.fa.gz")

if (!file.exists("ecoli_cds.fa")) {
  R.utils::gunzip("ecoli_cds.fa.gz")
}

list.files()
```
```{r}
#Download and Process C. coli CDS
#Campylobacter coli (GCA_003780985) CDS file
URL="https://ftp.ensemblgenomes.ebi.ac.uk/pub/bacteria/release-62/fasta/bacteria_46_collection/campylobacter_coli_gca_003780985/cds/Campylobacter_coli_gca_003780985.PDT000395653.1.cds.all.fa.gz"
download.file(URL,destfile="ccoli_cds.fa.gz")

if (!file.exists("ccoli_cds.fa")) {
  R.utils::gunzip("ccoli_cds.fa.gz")
}

list.files()
```


##Calculates the total number of Coding DNA Sequences (CDS) for both organisms
```{r}
library(Biostrings)
```

# Load the CDS files 
```{r}
ecoli_cds_dna <- readDNAStringSet("ecoli_cds.fa")
ccoli_cds_dna <- readDNAStringSet("ccoli_cds.fa")
```

# Calculate the number of coding sequences
```{r}
num_ecoli <- length(ecoli_cds_dna)
num_ccoli <- length(ccoli_cds_dna)
```


# Create the results table
```{r}
cds_count_table <- data.frame(
  Organism = c("E. coli (K-12 MG1655)", "C. coli (GCA_003780985)"),
  `Number of Coding Sequences` = c(num_ecoli, num_ccoli)
)
```

# Print the table
```{r}
print(cds_count_table)
```

##The main difference lies in their gene set size, which reflects their lifestyles: E.coli (with >2.7× the genes) is a versatile generalist, while C.coli is a specialized pathogen with a streamlined, reduced genome optimized for a stable host environment.



##Question 2
##Calculation of the total base pair length of all coding sequences 

```{r}
library(Biostrings)
```

# Load the CDS files 
```{r}
ecoli_cds_dna <- readDNAStringSet("ecoli_cds.fa")
ccoli_cds_dna <- readDNAStringSet("ccoli_cds.fa")
```

# Calculate the total length of coding DNA (sum of all sequence widths)
```{r}
total_length_ecoli <- sum(width(ecoli_cds_dna))
total_length_ccoli <- sum(width(ccoli_cds_dna))
```

# Create the results table using base R data.frame()
```{r}
total_length_table <- data.frame(
  Organism = c("E. coli (K-12 MG1655)", "C. coli (GCA_003780985)"),
  `Total_Coding_DNA_Length_(base_pairs)` = c(total_length_ecoli, total_length_ccoli),
  row.names = NULL
)
```

# Print the table
```{r}
print(total_length_table)
```
##E. coli possesses approx 2.75 times more total coding DNA, supporting its versatile generalist role with a higher gene count. Conversely, C. coli's reduced coding length reflects its streamlined, specialized pathogenic existence.

##Question 3
##Calculation of the mean and median lengths of all coding sequences 
##Generate a boxplot for visual comparison.

```{r}
library(Biostrings)
library(ggplot2)
```

##Load the CDS files
```{r}
ecoli_cds_dna <- readDNAStringSet("ecoli_cds.fa")
ccoli_cds_dna <- readDNAStringSet("ccoli_cds.fa")
```

##Calculate lengths for all sequences
```{r}
ecoli_lengths <- width(ecoli_cds_dna)
ccoli_lengths <- width(ccoli_cds_dna)
```

##Calculate Mean and Median
```{r}
ecoli_stats <- c(Mean = mean(ecoli_lengths), Median = median(ecoli_lengths))
ccoli_stats <- c(Mean = mean(ccoli_lengths), Median = median(ccoli_lengths))
```

##Create the summary table
```{r}
stats_df <- data.frame(
  Organism = c("E. coli (K-12 MG1655)", "C. coli (GCA_003780985)"),
  Mean_Length_bp = c(ecoli_stats["Mean"], ccoli_stats["Mean"]),
  Median_Length_bp = c(ecoli_stats["Median"], ccoli_stats["Median"]),
  row.names = NULL
)
print(stats_df)
```


##Create a combined data frame for the boxplot 
```{r}
length_df <- data.frame(
  Length = c(ecoli_lengths, ccoli_lengths),
  Organism = factor(c(rep("E. coli", length(ecoli_lengths)), rep("C. coli", length(ccoli_lengths))))
)
```


##Generate the Boxplot
```{r}
p_boxplot <- ggplot(length_df, aes(x = Organism, y = Length, fill = Organism)) +
  geom_boxplot() +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "Distribution of Coding Sequence Lengths",
    y = "Length (base pairs)",
    x = "Organism"
  ) +
  theme_minimal()
print(p_boxplot)
```
##The key difference lies not in average length (mean/median are similar), but in gene count and complexity. Both distributions are right-skewed, yet E. coli exhibits slightly more long gene outliers, reflecting its broader functional complexity and capacity for multi-domain proteins compared to the streamlined C. coli genome.


##Question 4
##Calculations and plots the frequency of DNA bases and amino acids for the total coding and protein sequences of both organisms.


##Nucleotide Frequency

##Calculate total length
```{r}
total_length_ecoli <- sum(width(ecoli_cds_dna))
total_length_ccoli <- sum(width(ccoli_cds_dna))
```

##Collapse all CDS into a single string for frequency calculation
```{r}
ecoli_nuc_freq <- alphabetFrequency(DNAStringSet(paste(ecoli_cds_dna, collapse = "")))[1, 1:4] / total_length_ecoli
ccoli_nuc_freq <- alphabetFrequency(DNAStringSet(paste(ccoli_cds_dna, collapse = "")))[1, 1:4] / total_length_ccoli
```

##Create data frame for plotting
```{r}
nuc_df <- data.frame(
  Nucleotide = names(ecoli_nuc_freq),
  Ecoli = ecoli_nuc_freq,
  Ccoli = ccoli_nuc_freq
)

nuc_df_long <- data.frame(
  Nucleotide = rep(nuc_df$Nucleotide, 2),
  Organism = c(rep("E. coli", 4), rep("C. coli", 4)),
  Frequency = c(nuc_df$Ecoli, nuc_df$Ccoli)
)
```


##Nucleotide Bar Plot
```{r}
p_nuc_bar <- ggplot(nuc_df_long, aes(x = Nucleotide, y = Frequency, fill = Organism)) +
  geom_bar(stat = "identity", position = position_dodge()) +
  labs(
    title = "Nucleotide Frequency in Total CDS", 
    y = "Proportion",
    subtitle = paste("E. coli G+C:", round(sum(ecoli_nuc_freq[c("G", "C")])*100, 1), 
                     "%, C. coli G+C:", round(sum(ccoli_nuc_freq[c("G", "C")])*100, 1), "%")
    ) +
  theme_minimal()
print(p_nuc_bar)
```



##Amino Acid Frequency

##Translate DNA to Protein Sequences
```{r}
ecoli_protein <- Biostrings::translate(ecoli_cds_dna)
ccoli_protein <- Biostrings::translate(ccoli_cds_dna)
```


##Collapse protein sequences and calculate frequency (using AA_ALPHABET to exclude symbols like '*')
```{r}
ecoli_aa_freq <- alphabetFrequency(AAStringSet(paste(ecoli_protein, collapse = "")))[1, AA_ALPHABET]
ccoli_aa_freq <- alphabetFrequency(AAStringSet(paste(ccoli_protein, collapse = "")))[1, AA_ALPHABET]
```

# Normalize
```{r}
ecoli_aa_freq_norm <- ecoli_aa_freq / sum(ecoli_aa_freq)
ccoli_aa_freq_norm <- ccoli_aa_freq / sum(ccoli_aa_freq)
```

# Create data frame for plotting
```{r}
aa_df_long <- data.frame(
  Amino_Acid = rep(names(ecoli_aa_freq_norm), 2),
  Organism = c(rep("E. coli", length(AA_ALPHABET)), rep("C. coli", length(AA_ALPHABET))),
  Frequency = c(ecoli_aa_freq_norm, ccoli_aa_freq_norm)
)
```


# Amino Acid Bar Plot
```{r}
p_aa_bar <- ggplot(aa_df_long, aes(x = Amino_Acid, y = Frequency, fill = Organism)) +
  geom_bar(stat = "identity", position = position_dodge()) +
  labs(title = "Amino Acid Frequency in Total Protein Sequence", y = "Proportion") +
  theme_minimal()
print(p_aa_bar)

```
##Nucleotide:C. coli is highly A/T-rich G+C approx 30%, imposing a major compositional bias on its coding sequences, whereas E. coli maintains a balanced G+C content near 50%.

##Amino Acid: This bias translates directly to protein: C. coli favors A/T -rich amino acids (e.g., Lysine, Isoleucine), demonstrating a genomic constraint where the nucleotide pool limits the available protein components; E. coli favors G/C-rich amino acids (e.g., Alanine, Glycine).

##Question 5
##Read coding DNA Sequences
```{r}
library(seqinr)
cds_ecoli <- read.fasta("ecoli_cds.fa")
cds_ccoli <- read.fasta("ccoli_cds.fa")
```
##Calculate RSCU
```{r}
codon_ecoli <- uco(unlist(cds_ecoli), index = "rscu", as.data.frame = TRUE)
codon_ccoli <- uco(unlist(cds_ccoli), index = "rscu", as.data.frame = TRUE)
```

##Sort codons by RSCU values

##E.coli
```{r}
sorted_ecoli <- codon_ecoli[order(codon_ecoli$RSCU), ]
top10_codons_ecoli_under <- head(sorted_ecoli, 10)
top10_codons_ecoli_over <- head(sorted_ecoli[order(-sorted_ecoli$RSCU), ], 10)
```

##C.coli
```{r}
sorted_ccoli <- codon_ccoli[order(codon_ccoli$RSCU), ]
top10_codons_ccoli_under <- head(sorted_ccoli, 10)
top10_codons_ccoli_over <- head(sorted_ccoli[order(-sorted_ccoli$RSCU), ], 10)
```

##Print top codons
```{r}
print(top10_codons_ecoli_under)
print(top10_codons_ecoli_over)
print(top10_codons_ccoli_under)
print(top10_codons_ccoli_over)
```


##Create Barplots
```{r}
par(mfrow = c(2, 2))  # 2x2 layout for four plots

barplot(top10_codons_ecoli_under$RSCU, names.arg = top10_codons_ecoli_under$codon,
        xlab = "Codons", ylab = "RSCU", col = "lightblue",
        main = "E. coli: 10 Least Used Codons", las = 2)

barplot(top10_codons_ecoli_over$RSCU, names.arg = top10_codons_ecoli_over$codon,
        xlab = "Codons", ylab = "RSCU", col = "steelblue",
        main = "E. coli: 10 Most Used Codons", las = 2)

barplot(top10_codons_ccoli_under$RSCU, names.arg = top10_codons_ccoli_under$codon,
        xlab = "Codons", ylab = "RSCU", col = "lightpink",
        main = "C. coli: 10 Least Used Codons", las = 2)

barplot(top10_codons_ccoli_over$RSCU, names.arg = top10_codons_ccoli_over$codon,
        xlab = "Codons", ylab = "RSCU", col = "tomato",
        main = "C. coli: 10 Most Used Codons", las = 2)
```

##Figure:Codon usage bias (RSCU) in Escherichia coli and Campylobacter coli.  
The plots show the top 10 over-represented and under-represented codons in each organism.  
 

E. coli shows a stronger codon bias, frequently using codons ending with G or C, consistent with its higher GC content and efficient translational machinery.  
Campylobacter coli exhibits weaker bias, with higher preference for A- or T-ending codons, reflecting its lower genomic GC content.  
 
These differences in codon bias may relate to variations in genomic composition, tRNA abundance, and adaptation to different environmental or metabolic conditions.

##Question 6
```{r}
library(Biostrings) 
library(kmer) 
library(dplyr) 
library(ggplot2)
```


##Load CDS files (using the Biostrings function, which is often cleaner)

```{r}
cds_ecoli <- readDNAStringSet("ecoli_cds.fa")
cds_ccoli <- readDNAStringSet("ccoli_cds.fa")
```

##Create protein object
```{r}
prot_ecoli <- Biostrings::translate(cds_ecoli)
prot_ccoli <- Biostrings::translate(cds_ccoli)
```

##Convert to character vectors
```{r}
prot_ecoli_vec <- as.character(prot_ecoli)
prot_ccoli_vec <- as.character(prot_ccoli)
```


##Define amino acid alphabet (excluding stop codons)
```{r}
aa <- sort(unique(unlist(strsplit(paste0(prot_ecoli_vec, collapse = ""), ""))))
aa <- aa[aa != "*"]
```

##Function to count protein k-mers
```{r}
# --- Function to count protein k-mers ---
get_protein_kmer_freq <- function(protein_seq, k) {
  # Collapse all sequences into one string
  all_seq <- paste(as.character(protein_seq), collapse = "")
  
  # Extract k-mers
  kmers <- substring(all_seq, 1:(nchar(all_seq)-k+1), k:(nchar(all_seq)))
  
  # Count frequency
  freq_table <- table(kmers)
  
  # Convert to data frame
  df <- as.data.frame(freq_table, stringsAsFactors = FALSE)
  colnames(df) <- c("kmer", "frequency")
  
  # Sort by descending frequency
  df <- df %>% arrange(desc(frequency))
  return(df)
}
```

##Compute 3-mer frequencies

```{r}
freq3_ecoli <- get_protein_kmer_freq(prot_ecoli, 3)
freq3_ccoli <- get_protein_kmer_freq(prot_ccoli, 3)
```

##Identify top 10 over- and under-represented k-mers
```{r}
over_ecoli <- head(freq3_ecoli, 10)
under_ecoli <- tail(freq3_ecoli, 10)
```

```{r}
over_ccoli <- head(freq3_ccoli, 10)
under_ccoli <- tail(freq3_ccoli, 10)
```

##Add organism column
```{r}
over_ecoli$Organism <- "E. coli"
under_ecoli$Organism <- "E. coli"
over_ccoli$Organism <- "C. coli"
under_ccoli$Organism <- "C. coli"
```

##Combine data frames for plotting
```{r}
over_df <- rbind(over_ecoli, over_ccoli)
under_df <- rbind(under_ecoli, under_ccoli)
```

##Plot over-represented 3-mers
```{r}
ggplot(over_df, aes(x = reorder(kmer, frequency), y = frequency, fill = Organism)) +
  geom_col(position = "dodge") +
  coord_flip() +
  labs(title = "Top 10 Over-represented 3-mers in Protein Sequences",
       x = "3-mer",
       y = "Frequency") +
  theme_minimal()
```

##Plot under-represented 3-mers
```{r}
ggplot(under_df, aes(x = reorder(kmer, frequency), y = frequency, fill = Organism)) +
  geom_col(position = "dodge") +
  coord_flip() +
  labs(title = "Top 10 Under-represented 3-mers in Protein Sequences",
       x = "3-mer",
       y = "Frequency") +
  theme_minimal()
```

